Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida

### 2021-12-08 manu-draft.Rmd
abstract shorten again, added highlights

From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA). This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days. Elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay. Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*. Reported fish kills tracked *K. brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish. Seagrass resources showed minimal changes over the study period.  By comparing these results to decades of baseline environmental monitoring data, we conclude that many of the biological responses, observed after the release from Piney Point, are abnormal relative to historic conditions.

>>>>

From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA). This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days. Elevated levels of phytoplankton (non-harmful diatoms) were first observed in April in the lower Bay. Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*. Reported fish kills tracked *K. brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish. Seagrasses had minimal changes over the study period.  By comparing these results to baseline environmental monitoring data, we conclude that many of the biological responses observed after the release from Piney Point are abnormal relative to historic conditions.
